Properties of winning Iterated Prisoner's Dilemma strategies.

### 2020-04-09 paper/main.tex
minor re writes by v.k.
Co-Authored-By: Vince Knight <vincent.knight@gmail.com>

Conceptualising strategies and understanding the best way of playing the game
has been of interest to the scientific community since the formulation of the
game in 1950~\cite{Flood1958}. Following the computer tournaments of Axelrod in the
1980's~\cite{Axelrod1980a, Axelrod1980b}, a strategy's performance in a round
robin computer tournament became a common evaluation technique for newly designed
strategies. Many tournaments have followed Axelrod's~\cite{Bendor1991, Stephens2002,
Kendall2007, Stewart2012, Harper2017} and today more hundreds of strategies exist
in the literature.

Axelrod performed two widely-known computer tournaments~\cite{Axelrod1980a,
Axelrod1980b} in the 1980's. The
winner of both tournaments was the simple strategy Tit For Tat (TFT) which cooperated
on the first turn and then simply copied the previous action of its opponent,
retailiating against defections with a defection, and forgiving a defection if followed
by a cooperation. Axelrod concluded that the strategy's robustness was due to four
properties, which he adapted into four suggestions on doing well in an IPD:

@@

There are strategies which have built upon TFT and the reciprocity based approach.
In~\cite{Beaufils1997} Gradual was introduced which was constructed to have the
same qualities as those of TFT except one: Gradual had a memory
of the game since the beginning of it. Gradual recorded the number of defections
by the opponent and punished them with a growing number of defections. It would
then enter a calming state in which it would cooperates for two rounds. A
strategy with the same intuition as Gradual is Adaptive Tit for
Tat~\cite{tzafestas-2000a}. Adaptive Tit for Tat does not keep a permanent
count of past defections, it maintains a continually updated estimate of the
opponent's behaviour, and uses this estimate to condition its future actions.

Other works have built upon the limitations of TFT, and others have shown
that suggestions made by Axelrod did not necessarily lift to alternative environmental rules.
In~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984} it was shown
that TFT suffered in environments with noise. This was
mainly due to the strategy's lack of generosity and contrition. Since TFT immediately
punishes a defection, in a noisy environment it can get stuck in a
repeated cycle of defections and cooperations. Some new strategies, more
robust in tournaments with noise, were soon introduced, including
Nice and Forgiving~\cite{Bendor1991}, Generous Tit For Tat~\cite{Nowak1992},
and Pavlov (aka Win Stay Lose Shift)~\cite{Nowak1993}, as well as later
variants such as OmegaTFT \cite{kendall2007iterated}. Introduction of new strategies
is often accompanied by a claim that the new strategy is the best known despite
only being tested against a small number of opponents, or specific classes of
opponents not necessarily representative of all possible or all published strategies.
The lack of testing against formally defined strategies and tournament winners
is understandable given the effort required to implement
\footnote{Implementing prior strategies faithfully is often extremely difficult or impossible
due to insufficient descriptions and lack of published implementations or code.}
the hundreds of published IPD strategies, yet calls into question any claims of superiority
or robustness of newly introduced
strategies.

A set of envious IPD strategies were introduced called zero-determinant strategies (ZDs)
in~\cite{Press2012}. By
forcing a linear relationship between stationary payoffs ZDs can ensure that they will
never receive less than their memory-one opponents. While ZDs
were introduced with a small tournament in which some were reportedly
successful \cite{Stewart2012}, this result has not generally held in future
work. In~\cite{Harper2017} a
series of clever strategies were introduced, and a tournament containing over
200 strategies featured no ZD strategies ranked in top
spots. Instead, the top ranked strategies were a set of ``clever''
(in the sense of Axelrod's characteristics) trained
strategies based on lookup tables~\cite{Axelrod1987}, hidden Markov
models~\cite{Harper2017}, and finite state automata~\cite{Miller1996}.
Similarly, in \cite{mathieu2017}, a set of evolutionarily-trained strategies,
and a pre-selected set of known strategies, outperformed a collection of 12 ZDs.

@@

In standard tournaments 10 out of the 15 top strategies are introduced
in~\cite{Harper2017}. These are strategies based on finite state automata (FSM),
hidden markov models (HMM), artificial neural networks (ANN), lookup tables
(LookerUp) and stochastic lookup tables (Gambler) that have been trained using
reinforcement learning algorithms (evolutionary and particle swarm algorithms).
They have been trained to perform well against the strategies
in APL in a standard tournament, thus their performance in the
specific setting was anticipated. DoubleCrosser, BackStabber and Fool Me Once, are
strategies not from the literature but from the APL. DoubleCrosser is an extension
of BackStabber and both strategies make use of the number of turns because they are
set to defect on the last two rounds. It should be noted that these
strategies can be characterised as ``cheaters'' because the source code of the strategies
allows them to know the number of turns in a match (if they are specified).
These strategies were expected to not perform as well in
tournaments where the number of turns is not specified. Finally, Winner
12~\cite{mathieu2017} and DBS~\cite{Au2006} are both from the literature.
DBS is a strategy specifically designed for noisy environments, however, it ranks
highly in standard tournaments as well. Similarly the fourth ranked player,
Evolved FSM 16 Noise 05, was
trained for noisy tournaments yet performs well in standard tournaments.
Figure~\ref{fig:std_results} shows that these strategies typically perform
well in any standard tournament in which they participate.

@@

In comparison, the performance of the top ranked strategies in noisy environments
when \(p_n\in [0, 1]\) is bimodal. The top strategies include strategies which
decide their actions based on the cooperation to defection ratio, such as
ShortMem~\cite{Andre2013}, Grumpy~\cite{axelrodproject} and
e~\cite{axelrodproject}, and the Retaliate strategies which are designed to
defect if the opponent has tricked them more often than x\% of the times that
they have done the same. The bimodality of the \(r\) distributions is explained
by Figure~\ref{fig:effect_of_noise} which demonstrates that the top 6 strategies
were highly ranked due to the their performance in tournaments with \(p_n>0.5\),
and that in tournaments with a noise probability lower than \(0.5\) they
performed poorly. At a noisy level of \(0.5\) or greater, mostly cooperative strategies
become mostly defectors and vice versa.

>>>>

Conceptualising strategies and understanding the best way of playing the game
has been of interest to the scientific community since the formulation of the
game in 1950~\cite{Flood1958}. Following the computer tournaments of Axelrod in the
1980's~\cite{Axelrod1980a, Axelrod1980b}, a strategy's performance in a round
robin computer tournament became a common evaluation technique for newly designed
strategies. Many tournaments have followed Axelrod's~\cite{Bendor1991, Beaufils1997, Stephens2002,
Kendall2007, Stewart2012, Harper2017} and today more hundreds of strategies exist
in the literature.

The
winner of both of Axelrod's tournaments~\cite{Axelrod1980a,
Axelrod1980b} was the simple strategy Tit For Tat (TFT) which cooperated
on the first turn and then simply copied the previous action of its opponent,
retailiating against defections with a defection, and forgiving a defection if followed
by a cooperation. Axelrod concluded that the strategy's robustness was due to four
properties, which he adapted into four suggestions on doing well in an IPD:

@@

There are strategies which have built upon TFT and the reciprocity based approach.
In~\cite{Beaufils1997} Gradual was introduced which was constructed to have the
same qualities as those of TFT except one: Gradual had a memory
of the game since the beginning of it. Gradual recorded the number of defections
by the opponent and punished them with a growing number of defections. It would
then enter a calming state in which it would cooperates for two rounds. A
strategy with the same intuition as Gradual is Adaptive Tit for
Tat~\cite{tzafestas-2000a}. Adaptive Tit for Tat maintains a continually updated estimate of the
opponent's behaviour, and uses this estimate to condition its future actions.

Other works have built upon the limitations of TFT, and others have shown
that suggestions made by Axelrod did not necessarily apply in alternative environmental settings.
In~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984} it was shown
that TFT suffered in environments with noise. This was
mainly due to the strategy's lack of generosity and contrition. Since TFT immediately
punishes a defection, in a noisy environment it can get stuck in a
repeated cycle of defections and cooperations. Some new strategies, more
robust in tournaments with noise, were soon introduced, including
Nice and Forgiving~\cite{Bendor1991}, Generous Tit For Tat~\cite{Nowak1992},
and Pavlov (aka Win Stay Lose Shift)~\cite{Nowak1993}, as well as later
variants such as OmegaTFT \cite{kendall2007iterated}. Introduction of new strategies
is often accompanied by a claim that the new strategy is the best known despite
only being tested against a small number of opponents, or specific classes of
opponents not necessarily representative of all possible or all published strategies.
The lack of testing against formally defined strategies and tournament winners
is understandable given the effort required to implement
\footnote{Implementing prior strategies faithfully is often extremely difficult or impossible
due to insufficient descriptions and lack of published implementations or code.}
the hundreds of published IPD strategies, yet calls into question any claims of superiority
or robustness of newly introduced
strategies.

A set of envious IPD strategies were introduced called zero-determinant strategies (ZDs)
in~\cite{Press2012}. By
forcing a linear relationship between stationary payoffs ZDs can ensure that they will
never receive less than their memory-one opponents. While ZDs
were introduced with a small tournament in which some were reportedly
successful \cite{Stewart2012}, this result has not generally held in future
work. In~\cite{Harper2017} a
series of strategies trained using reinforcement learning were introduced, and a tournament containing over
200 strategies featured no ZD strategies ranked in top
spots. Instead, the top ranked strategies were a set of ``clever''
(in the sense of Axelrod's characteristics) trained
strategies based on lookup tables~\cite{Axelrod1987}, hidden Markov
models~\cite{Harper2017}, and finite state automata~\cite{Miller1996}.
Similarly, in \cite{mathieu2017}, a set of evolutionarily-trained strategies,
and a pre-selected set of known strategies, outperformed a collection of 12 ZDs.

@@

In standard tournaments 10 out of the 15 top strategies are introduced
in~\cite{Harper2017}. These are strategies based on finite state automata (FSM),
hidden markov models (HMM), artificial neural networks (ANN), lookup tables
(LookerUp) and stochastic lookup tables (Gambler) that have been trained using
reinforcement learning algorithms (evolutionary and particle swarm algorithms).
They have been trained to perform well against the strategies
in APL in a standard tournament, thus their performance in the
specific setting was anticipated. DoubleCrosser, BackStabber and Fool Me Once, are
strategies not from the literature but from the APL. DoubleCrosser is an extension
of BackStabber and both strategies make use of the number of turns because they are
set to defect on the last two rounds. It should be noted that these
strategies can be characterised as ``cheaters'' because the source code of the strategies
allows them to know the number of turns in a match (unless the match has a probabilistic ending).
These strategies were expected to not perform as well in
tournaments where the number of turns is not specified. Finally, Winner
12~\cite{mathieu2017} and DBS~\cite{Au2006} are both from the literature.
DBS is a strategy specifically designed for noisy environments, however, it ranks
highly in standard tournaments as well. Similarly the fourth ranked player,
Evolved FSM 16 Noise 05, was
trained for noisy tournaments yet performs well in standard tournaments.
Figure~\ref{fig:std_results} shows that these strategies typically perform
well in any standard tournament in which they participate.

@@

In comparison, the performance of the top ranked strategies in noisy environments
when \(p_n\in [0, 1]\) is bimodal. The top strategies include strategies which
decide their actions based on the cooperation to defection ratio, such as
ShortMem~\cite{Andre2013}, Grumpy~\cite{axelrodproject} and
e~\cite{axelrodproject}, and the Retaliate strategies which are designed to
defect if the opponent has tricked them more often than a given percentage of the times that
they have done the same. The bimodality of the \(r\) distributions is explained
by Figure~\ref{fig:effect_of_noise} which demonstrates that the top 6 strategies
were highly ranked due to the their performance in tournaments with \(p_n>0.5\),
and that in tournaments with a noise probability lower than \(0.5\) they
performed poorly. At a noisy level of \(0.5\) or greater, mostly cooperative strategies
become mostly defectors and vice versa.
